---
title: "Reading Guide"
author: "Dan Maxwell"
date: "May 3, 2017"
output: pdf_document 
header-includes: \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section Notes for *R in Action*

**1.2** Obtaining and Installing R  
You must install base R before installing RStudio.  In fact, RStudio won't work without base R installed. \newline

**1.3** Working with R  
In this section, Kabacoff assumes that the reader already knows what a function is and how they are called.  As defined by Webopedia, a software function is "a named section of a program that performs a specific task."  By now, you should have already read the section on functions, written by Peter Dalgaard.  A link to that document is provided in Canvas. \newline

**1.3.1** Getting Started  
Because we are using the RStudio integrated development environment (IDE), what you see in the RStudio console will differ from Figure 1.3. \newline 

**1.3.3** The workspace  
For newcomers, the `getwd ()` and `setwd ()` functions are critically important because R sources files from your current working directory.  Inevitably, students forget where they're at and then get an error message when trying to load a data file. \newline

```{r}
setwd("c:/Informatics")  # Set the working directory -- This is now the default directory
getwd()                  # Get the working directory -- Where are we at?
```

**1.7** Working with large datasets  
Appendix F is worth the read, especially if you think you'll be working with very large datasets.  Memory management in R is unique in that function arguments are passed by *value* and not by *reference*.  What this means, in practical terms, is that R makes a copy of each object you pass to a function.  So if you pass a dataframe to a function, you now have two copies of that dataframe in memory, one for the function and another for the program that called it.  Because of this, R will gobble up a lot of memory when your datasets (dataframes) are really large. \newline


