--- 
title: | 
       | *A Picture is Worth a Thousand Words*
       | A myDataStory Learning Experience
author: "Dan Maxwell"
date: "August 12, 2018"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{color}
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)  # FALSE hides all code...
```


```{r init, echo = FALSE, message = FALSE}

library(dplyr)

```

## Introduction
Up to this point, our focus has been data management.  Newcomers to R often want to rush through data scrubbing activities as they find this work tedious.  They would much rather spend their time creating sensational graphs of their data as visualization work is a lot more fun.  However, most data scientists will tell you that the bulk of their time (80% or more) is spent preparing the data.  Indeed, creating a first graph often reveals additional data issues, an outlier that needs to be handled or values that need to be imputed.  Often a visualization/data cycle sets up, and you'll find yourself going back and forth between looking at the data and fixing it.  Eventually, the data "settles down" and the graphs begin to tell a story.  Keep this in mind as you work through the examples provided here. 

## A Helpful Resource
See http://www.statmethods.net/graphs/index.html for additional graphing resources.

## The Basic Histogram

```{r basic_histogram}
# Generate random numbers from a normal distribution.
x <- rnorm(mean = 5, sd = 1, n = 50)

# Create a histogram with a title, indicating the mean and standard deviation.
hist(x, main = expression(paste("Sampled values, ", mu, " = 5, ", sigma, " = 1")),
     col = "lightblue")
```

## The Basic Plot

```{r basic_plot}

# Create dataframe from the data file.
data <- read.csv("https://github.com/mydatastory/r_intro_class/raw/master/data/gapminder.csv")

# Create a subset of data for just the United States.
us  <- data[data$country=="United States",]

# Plot GDP per Capital by year.
plot(us$year, us$gdpPercap,
     xlab = "Year",
     ylab = "GDP per Capital",
     type = "b", 
     cex  = .75,
     lty  = 3,
     col  = "red",
     pch  = 1)

```


## The Multi-Line Plot

```{r multiline_plot}

# Create vectors, one for each country represented.
us <- data[data$country=="United States",]
uk <- data[data$country=="United Kingdom",]

x <- us$year
y <- us$gdpPercap

plot(x, y,
     xlab = "Year",
     ylab = "GDP per Capital",
     type = "b", 
     cex  = .75,
     lty  = 3,
     col  = "blue",
     pch  = 1,
     las  = 1,
     cex.axis = .60)

x <- uk$year
y <- uk$gdpPercap

# Now plot the UK line, using a different color and line type.
lines(x, y, type = "b", lty = 2, col = "red", pch = 0, cex = .75)

# Add legend to distinguish between the two lines.
legend("bottomright",
       title = "GDP Per Cap",
       c("United States","United Kingdom"),
       lty   = c(3, 2), 
       pch   = c(1, 0),
       col   = c("blue","red"))

```

Now add GDP lines for Germany and Japan to the graph...

## Advanced Histograms

```{r life_histograms}

# Create dataframe from data file.
life_exp_1987 <- data[data$year==1987,]
life_exp_2007 <- data[data$year==2007,]

# Set graphic device parameters to 2 rows and 1 column.
par(mfrow = c(2, 1))    

# Graph the two vectors.
hist(life_exp_1987$lifeExp, main = "Life Expectancy in 1987", 
     xlab = "Life Expectancy", xlim = range(30:90), col = "lightblue")
hist(life_exp_2007$lifeExp, main = "Life Expectancy in 2007", 
     xlab = "Life Expectancy", xlim = range(30:90), col = "lemonchiffon")

# Restore graphic device to 1 row, 1 column.
par(mfrow = c(1, 1))    

```


## Advanced Boxplots

```{r life_boxplots}

# Two ways to create the boxplot.
# Create with filter data
boxplot(life_exp_1987$lifeExp, life_exp_2007$lifeExp, 
        names = c("1987","2007"), col = "rosybrown1")

boxplot(data$lifeExp ~ data$year, 
        data[data$year==1987|2007,], col = "rosybrown1")
        
# Create with formula (all years)
boxplot(data$lifeExp ~ data$year, col = "rosybrown1")

```


### The Basic Barplot

```{r barplot}

## Basic barplot

set.seed(123)
rows_for_barplot <- data[sample(nrow(data), 6), ]

barplot(rows_for_barplot$lifeExp, 
        names.arg = paste0(rows_for_barplot$country, "\n", rows_for_barplot$year),
        cex.names = 0.8,
        col = rainbow(6, s = 0.3),
        ylab = "Life Exp (Years)",
        ylim = c(0,80))
```


## Grouped & Stacked Barplots

```{r group_barplot, echo = FALSE}

library(dplyr)

# Remove Oceania because it only has a sample size of 2 (New Zealand and Australia)
pop_2007 <- dplyr::filter(data, year == 2007, continent != "Oceania")
pop_2007 %>% group_by(continent) %>% summarise(sum(pop))

pop_2007 %>% group_by(continent) %>% mutate(cont_pop = cumsum(pop))

for (cont in unique(pop_2007$continent)){
  x <- dplyr::filter(pop_2007, continent == cont) %>% top_n(3, pop) %>% droplevels
  
  # Assign the name of the cont as the variable name
  assign(cont, x)
}

# Data must be presented as a matrix or vector for stacked or grouped barplots.
m1 <- cbind(Africa = Africa$pop, Americas = Americas$pop, 
            Asia = Asia$pop, Europe = Europe$pop)
m1

xx <- rep(0,3)
aa <- Africa$pop
bb <- Americas$pop
cc <- Asia$pop
dd <- Europe$pop
m2 <- matrix(c(aa,xx,xx,xx,
               xx,bb,xx,xx,
               xx,xx,cc,xx,
               xx,xx,xx,dd), ncol = 4)
colnames(m2) <- c("Africa", "Americas", "Asia", "Europe")

m2

barplot(m2, 
        legend = cbind(levels(Africa$country), levels(Americas$country), 
                       levels(Asia$country), levels(Europe$country)),
        col = rainbow(12, s = 0.5),
        beside = F)

barplot(m1, 
        legend = cbind(levels(Africa$country), levels(Americas$country), 
                       levels(Asia$country), levels(Europe$country)),
        col = rainbow(12, s = 0.5),
        beside = T)

```


## The Table Function

When you need to summarize a dataframe, the `table()` function is often useful. 

```{r continents}

continents <- data[,c("country","continent")]
continents <- unique(continents)
cont_table <- table(continents$continent)
cont_table

```


## The Pie Chart

```{r pie}

cont_table
lbls <- paste0(names(cont_table), "\n ", cont_table)
pie(cont_table, labels = lbls, main = "Pie Chart of Continents\n (number of countries)")

```


## The Scatter Plot

```{r scatterplot, echo = FALSE}

data_2007 <- data[data$year==2007,]
plot(data_2007$gdpPercap, data_2007$lifeExp, xlab = "GDP Per Capital ($)", ylab = "Life Expectancy (Years)")

```


## The Stem-and-Leaf Plot

```{r stemleaf, echo = FALSE}

# The scale argument must equal 2 or the function only prints even numbers.  This
# is not stated in the help documentation.  The answer was found on Stack Overflow.

stem(data_2007$lifeExp, scale = 2) 

```


## Mathematical Symbols in Plots
Now that we've demonstrated some of the graphs available to you in base R, there may be times when you'll want to add special mathematical characters to a plot, usually letters from the Greek alphabet.  In order to do that, you'll need to use the `expression()` function.  Let's take a look at some code to see how this is done: 

```{r, fig.width = 6.5, fig.height = 4.10}
# Create a blank plot.
plot(1:4, 1:4, type = "n")

# Add text inside the plot to demonstrate how to embed math symbols and/or equations.
text(1.5, 3, expression(mu))
text(2.5, 3, expression(sigma))
text(3.5, 3, expression(pi))
text(1.5, 2, expression(sqrt(x, y)))
text(2.5, 2, expression(frac(sum(x), n)))
text(3.5, 2, expression(x^2))

```

The code listed here works fine if all you want to do is embed a formula.  But
what if you also want to add explanatory text?  In that case, you'll need to use the `paste()` function.  Let's generate a simple histogram to show how this is done.

```{r}
# Generate random numbers from a normal distribution.
x <- rnorm(mean = 5, sd = 1, n = 50)

# Create a histogram with a title, indicating the mean and standard deviation.
hist(x, main = expression(paste("Sampled values, ", mu, " = 5, ", sigma, " = 1")),
         col = "lightblue")
```

We now have a nice title for our graph with embedded symbols.  R provides support for just about any kind of mathematical expression.  To see more examples, type `?plotmath()` at the R prompt.

## References  



 


