--- 
title: | 
       | Activities Worksheet
       | *Base R Plotting 1*
author: "Dan Maxwell"
date: "June 1, 2017"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{color}
---
 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)  # FALSE hides all code...
```


```{r init, echo = FALSE, message = FALSE}

library(dplyr)

```

## Introduction
Up to this point, our focus has been data management.  Newcomers to R often want to rush through data scrubbing activities as they find this work tedious.  They would much rather spend their time creating sensational graphs of their data as visualization work is a lot more fun.  However, most data scientists will tell you that the bulk of their time (80% or more) is spent preparing the data.  Indeed, creating a first graph often reveals additional data issues, an outlier that needs to be handled or an extreme data point that must be recoded.  Often a visualization/data cycle sets up, and you'll find yourself going back and forth between looking at the data and fixing it.  Eventually, the data "settles down" and the graphs begin to tell a story.  Keep this in mind as you work through the challenges in this worksheet. 

## Mathematical Symbols in Plots
Many times you need to add special mathematical characters to your graphs, usually letters from the Greek alphabet.  In order to do that, you'll need to use the `expression()` function.  Let's take a look at some code to see how this is done: 

```{r, fig.width = 6.5, fig.height = 4.10}
# Create a blank plot.
plot(1:4, 1:4, type = "n")

# Add text inside the plot to demonstrate how to embed math symbols and/or equations.
text(1.5, 3, expression(mu))
text(2.5, 3, expression(sigma))
text(3.5, 3, expression(pi))
text(1.5, 2, expression(sqrt(x, y)))
text(2.5, 2, expression(frac(sum(x), n)))
text(3.5, 2, expression(x^2))

```

The code listed here works fine if all you want to do is embed a formula.  But
what if you also want to add explanatory text?  In that case, you'll need to use
the `paste()` function.  Let's generate a simple histogram to show how this is
done.

```{r}
# Generate random numbers from a normal distribution.
x <- rnorm(mean = 5, sd = 1, n = 50)

# Create a histogram with a title, indicating the mean and standard deviation.
hist(x, main = expression(paste("Sampled values, ", mu, " = 5, ", sigma, " = 1")),
         col = "lightblue")
```

We now have a nice title for our graph with embedded symbols.  R provides
support for just about any kind of mathematical expression.  To see more
examples, type `?plotmath()` at the R prompt.

## Helpful Resource for the Challenges 
Robert Kabacoff, the author of *R in Action*, provides a quick reference guide of R graphs at the following website:

http://www.statmethods.net/graphs/index.html

## Challenge 1 (The Basic Plot)
For this first plot challenge, you will plot the percentage of families in poverty, from 2005 to 2015.  The file that contains this information is family_poverty.txt.  What effect did the great recession have on these percentages?  Your final graph should look something like this:    

```{r basic_plot}

# Create dataframe from the data file
data <- read.csv("../data/gapminder.csv")

us  <- subset(data, country == "United States")

# Plot GDP per Capital by year.
plot(us$year, us$gdpPercap,
               xlab = "Year",
               ylab = "GDP per Capital",
               type = "b", 
               cex  = .75,
               lty  = 3,
               col  = "red",
               pch  = 1)


```
Data Source: *U.S. Census Bureau*
 

## Challenge 2 (Multi-Line Plot)
In this second plot challenge, load the race_poverty.txt file into a dataframe and then plot the poverty numbers for whites and hispanics.  As this dataset also contains poverty numbers for asians and blacks, you could also create a graphic that plots all four.  In either case, the final plot should look something like this:

```{r multiline_plot}

# Create vectors, one for each country represented.
uk         <- subset(data, country == "United Kingdom")
germany    <- subset(data, country == "Germany")
japan      <- subset(data, country == "Japan")

x <- us$year
y <- us$gdpPercap

plot(x, y,
     xlab = "Year",
     ylab = "GDP per Capital",
     type = "b", 
     cex  = .75,
     lty  = 3,
     col  = "blue",
     pch  = 1,
     las  = 1,
     cex.axis = .60)

x <- uk$year
y <- uk$gdpPercap

# Now plot the UK line, using a different color and line type.
lines(x, y, type = "b", lty = 2, col = "red", pch = 0, cex = .75)

# Add legend to distinguish between the two lines.
legend("bottomright",
       title = "Race",
       c("United States","United Kingdom"),
       lty   = c(3, 2), 
       pch   = c(1, 0),
       col   = c("blue","red"))

```
Data Source: *U.S. Census Bureau*

 
## Challenge 3 (The Histogram and Boxplot)

In this last challenge, you're going to compare energy expenditures for lean and obese women.  The data file is energy_expend.csv.  First create two histograms to compare the distributions of these two groups, and then create a boxplot and four stripcharts.  For  the stripcharts, first create a basic graph and then ones where the method argument equals "stack" and "jitter."  Your graphs should look like those pictured here.  From looking at these graphs, What conclusions can be drawn about these two groups?

```{r energy_histogram}

# Create dataframe from data file.
life_exp_1987 <- dplyr::filter(data, year==1987)
life_exp_2007 <- dplyr::filter(data, year==2007)

# Create two vectors, one for each stature factor.


# Set graphic device parameters to 2 rows and 1 column.
par(mfrow = c(2, 1))    

# Graph the two vectors.
hist(life_exp_1987$lifeExp, main = "Life Expectancy in 1987", xlab = "Life Expectancy", xlim = range(30:90), col = "lightblue")
hist(life_exp_2007$lifeExp, main = "Life Expectancy in 2007", xlab = "Life Expectancy", xlim = range(30:90), col = "lemonchiffon")

# Restore graphic device to 1 row, 1 column.
par(mfrow = c(1, 1))    

```

Now, let's create a boxplot of the data...


```{r energy_boxplot}

# Two ways to create the boxplot.
# Create with filter data
boxplot(life_exp_1987$lifeExp, life_exp_2007$lifeExp, names = c("1987","2007"), col = "rosybrown1")
boxplot(data$lifeExp ~ data$year, subset = data$year %in% c(1987,2007), col = "rosybrown1")

# Create with formula (all years)
boxplot(data$lifeExp ~ data$year, col = "rosybrown1")

```
Data Source: *Introductory Statistics with R* (p. 79 - 83)


#### References  

Dalgaard, P. (2008). *Introductory statistics with R* (2nd ed.). New York: Springer.

Census Bureau: https://www.census.gov/data/tables/time-series/demo/income-poverty/historical-poverty-people.html

 


