---
title: "Exercise Worksheet (Chapter 6)"
author: "Dan Maxwell"
date: "June 1, 2017"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r init, echo = FALSE, warning = FALSE, message = FALSE}

library(sqldf)
library(dplyr)

```

## Introduction
Review the section notes for *R in Action* before working on the challenges.

> ### *Ref: R in Action (Section 6.1.1)* 
> As Kabacoff does not discuss the `table()` function here, you'll want to read
> about it in section 7.2.1 (pages 145 & 146). 

> ### *Ref: R in Action (Section 6.5.1)* 
> The `~` in the formula `mpg ~ cyl` separates the *dependent* or *response*
> variable on the left from the *independent* (*explanatory* or *predictor*)
> variable on the right.  Please note: statistics textbookss are inconsistent in
> the names they use for these two variables.    
>
> Here's a definition of these two kinds of variables, provided by Gravetter &
> Wellnau (2016).  "The *dependent variable* is the one that is observed to assess
> the effect of the treatment ... while the *independent variable* is the one
> manipulated by the researcher" (p. 15).  
>
> In Listing 6.9, Kabacoff introduces the `*` operator. What does that mean? To
> answer this question, see table 8.2 (p. 172) for a list of symbols used in R
> formulas and their respective definitions. \newline


### Challenge 1 (The Basic Barplot)

In this first challenge, load the informatics_survey.csv file into a dataframe and then select
just the question 5 (Q5_1) column.  From there, create a basic barplot that looks like this:

```{r barplot}

survey <- read.csv("informatics_survey.csv", stringsAsFactors = FALSE)

# What happens if we don't set the stringsAsFactors argument?
q <- data.frame(survey$Q5_1, stringsAsFactors = FALSE)

colnames(q) <- c("response")

# Remove the rows with empty responses.
q <- subset(q, response != "")
q <- q[q$response != "",]

cnts <- table(q)

# Save the original graph parameters.
opar <- par(no.readonly = TRUE)

# Adjust graph margin parameters.
par(mai = c(.5, 1.5, .5, .5))

# What does the las argument do?  Look it up with ?par().
barplot(cnts, 
        horiz = TRUE, 
        col   = 'lightblue',
        las   = 2,
        main  = "Knowledge of Statistics - Question 5.1",
        cex.names = 0.5)

par(opar)

```
Data Source: *2017 UF Informatics Survey*

## Challenge 2 (The Grouped & Stacked Barplot)

In this second challenge, load the drink_crime.csv file into a dataframe and then generate both
a stacked and grouped barplot.  This drink_crime.csv file lists the percentage of criminal
activity in relation to one's drinking level.  Are regular drinkers more likely to enage in
criminal activity?  When completed, your graph should look something like this:


```{r group_barplot}

crime <- read.csv("drink_crime.csv", stringsAsFactors = FALSE)

# Data must be presented as a matrix or vector for stacked or grouped barplots.

barplot(as.matrix(crime[, c(2:4)]), 
        legend = crime[, 1],
           col = c("lightcoral","lightblue","khaki","beige"))

barplot(as.matrix(crime[, c(2:4)]), 
        legend = crime[, 1],
           col = c("lightcoral","lightblue","khaki","lightgreen"),
   args.legend = list(x = "topleft"),
        beside = TRUE)

# INSTRUCTOR NOTE
# See activities_chp_6.R for another way to solve this challenge.

```
Data Source: *Statistical Methods and Data Analysis* (p. 110)


## Challenge 3 (The Pie Chart)

In 2017, the Informatics Institue at the University of Florida conducted a survey to determine
the training needs of its clientele.  The informatics_survey.csv file contains the data from 
that survey.  In this challenge, you'll create a pie-chart that displays the percentage of
respondents who came from the Sciences, Health Sciences, Humanities, and Social Sciences. You will also need to create a slice for 'Unknown' as some respondents left this question blank.

Question 3 (q3) on the survey asked participants to select an academic department, but it does not classify those departments.  Fortunately, the informatics_acad_grp.csv file provides group totals by department and an acad_grp column with the following values: (UNK) Unknown, (SSI) Social Sciences, (HUM) Humanities, (SCI) Sciences, and (MED) Health Sciences.  You can use that file for this challenge or you can recreate it from the informatics_survey.csv file - a much bigger challenge that will require you to group and manually edit the data.  Please note: this advanced challenge is optional.

When completed, your graph should look something like this:

```{r pie}

# Solution for the optional (advanced) challenge to create the 
# informatics_acad_grp.csv file from the informatics_survey.csv file 
# rather than use the one provided.

survey <- read.csv("informatics_survey.csv", stringsAsFactors = FALSE)

# INSTRUCTOR NOTE 
# sqldf does not work when running under R version 3.4.0.  The two commands below
# return an error message that the survey table cannot be found, though it was
# created above.  As of 06.26.17, sqldf calls for R version 3.3.3 but works with
# warnings under 3.3.2.

# You can write a select statement with new column names. 
tmp <- sqldf("select q3 as department, count(q3) as total from survey where q3 != ' ' group by q3")

# Or, you can execute a simple select and rename the variables afterwards.
tmp <- sqldf("select q3, count(q3) from survey where q3 != ' ' group by q3")

# Rename the columns -- department & department count.
colnames(tmp)  <- c("dept","dept_cnt")

# Add a new column for academic group with the following values:
# (HUM Humanities, SCI Sciences, SOC, Social Sciences, MED Medical)
tmp$acad_grp <- ""                       

# INSTRUCTOR NOTE
# In order to populate the new acad_grp variable, students will need to edit the dataframe
# with the fix() function.  This opens a simple text editor which allows data entry 
# and modification.  Strangely, the edit() function opens the same editor but changes 
# are not permanently saved to the dataframe.

# Save a backup copy of the modified dataframe.
write.csv(tmp, file = "temp.csv")          

# Create a new dataframe of the two columns needed for the pie chart. 
temp   <- data.frame(tmp$acad_grp, tmp$dept_cnt, stringsAsFactors = FALSE)

colnames(temp) <- c("acad_grp","dept_cnt")

# End optional challenge -------------------------------------------------

temp <- read.csv("informatics_acad_grp.csv", stringsAsFactors = FALSE)

# Group rows by academic group.
grp_by <- group_by(temp, acad_grp)                    

# Sum dept_cnt for each academic group.
slices <- summarize(grp_by, grp_tot = sum(dept_cnt))

# Assign descriptive labels to the slices.
slices$acad_grp[1] <- "Humanities"
slices$acad_grp[2] <- "Health Sciences"
slices$acad_grp[3] <- "Sciences"
slices$acad_grp[4] <- "Social Sciences"
slices$acad_grp[5] <- "Unknown"

# Calculate percentages, round to two digits, and then create labels.
slices$grp_pct <- (slices$grp_tot / sum(slices$grp_tot))
slices$grp_pct <- round(slices$grp_pct, 2) * 100
slices$lbl     <- paste(slices$acad_grp, " ", slices$grp_pct, "%", sep = "")

pie(slices$grp_tot, labels = slices$lbl, main = "Pie Chart") 

# INSTRUCTOR NOTE
# This code accomplishes the same thing using the pipe operator.  If you have
# time to demo this in lab, uncomment and test.

# slices <- group_by(temp, acad_grp) %>% summarize(grp_tot = sum(dept_cnt))
# pie(slices$grp_tot, labels = slices$acad_grp, main = "Pie Chart (Pipe)") 

```
Data Source: *2017 UF Informatics Survey*
 

## Challenge 4 (The Histogram)

The home_ownership.txt file contains home ownership rates for all 50 states and the District of Columbia, for the years 1985, 1996, and 2002.  In this challenge, you will create three histograms, one for each year.  When complete, your graphs should look like these:

```{r histogram}

opar <- par(no.readonly = TRUE)

own <- read.delim("home_ownership.txt", sep = "\t", stringsAsFactors = FALSE)

hist(own$pct_1985, 
     main = "1985", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

hist(own$pct_1996, 
     main = "1996", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

hist(own$pct_2002, 
     main = "2002", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

par(opar)

```
Data Source: *Statistical Methods and Data Analysis* (p. 129)

## Challenge 5 (The Box Plot)

Now, create side-by-side boxplots of the home_ownership dataset used in the last challenge. Your graph should look like this when you're done:

```{r boxplot}

own <- read.delim("home_ownership.txt", sep = "\t", stringsAsFactors = FALSE)

boxplot(own$pct_1985, 
        own$pct_1996, 
        own$pct_2002, 
        notch = TRUE,
        col   = "tan",
        ylab  = "Home Ownership %",
        main  = "Home Ownership in the United States",
        names = c("1985","1996","2002"))

```
Data Source: *Statistical Methods and Data Analysis* (p. 129)

## Challenge 6 (The Scatter Plot)

In this challenge you will create a basic scatterplot to compare the M2 and M3 money supply (in trillions of dollars).  The data is found in money_supply.txt.  Also, fit a linear regression model to the data and plot the resulting line.  Your graph should look like this when complete:

```{r scatterplot}

money <- read.delim("money_supply.txt", sep = "\t", stringsAsFactors = FALSE)

plot(money$m2, money$m3, xlab = "M2 (Trillions)", ylab = "M3 (Trillions)")

# Now, fit a regression model and plot the line.
fit <- lm(money$m3 ~ money$m2, data = money)

abline(fit, col = "blue")

```
Data Source: *Statistical Methods and Data Analysis* (p. 137)


## Challenge 7 (The Stem-and-Leaf Plot)

In this last challenge, create a stem-and-leaf plot of the data found in ozone.csv.  According to Ott (2016), these "are the maximum ozone readings taken on 80 summer days in a large city" (p. 75).  Your graph should look like this when complete:  

```{r stemleaf}

ozone <- read.csv("ozone.csv", stringsAsFactors = FALSE)

x <- ozone[c(0:79),]

# INSTRUCTOR NOTE
# The scale argument must equal 2 or the function only prints even numbers.  This
# is not stated in the help documentation.  The answer was found on Stack Overflow.

stem(x, scale = 2) 

```
Data Source: *Statistical Methods and Data Analysis* (p. 77)


#### References

Gravetter, F. J. (2015). *Statistics for the behavioral sciences* (10th ed.). Boston, MA: Cengage Learning.

Ott, L., & Longnecker, M. (2016). *An introduction to statistical methods & data analysis* (7th ed.). Boston, MA: Cengage Learning.

