---
title: "Exercise Worksheet (Chapter 6)"
author: "Dan Maxwell"
date: "June 1, 2017"
output:
  pdf_document: default
  html_document: default
header-includes: \usepackage{color}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the datafile and supporting libraries...

```{r init}

library(sqldf)
library(dplyr)

```


### The Barplot Challenge

```{r barplot}

survey <- read.csv("informatics_survey.csv", stringsAsFactors = FALSE)

# What happens if we don't set the stringsAsFactors argument?
q5_1 <- data.frame(survey$Q5_1, stringsAsFactors = FALSE)

colnames(q5_1) <- c("response")

# Remove the rows with empty responses.
q5_1 <- subset(q5_1, response != "")
q5_1 <- q5_1[q5_1$response != "",]

cnts <- table(q5_1)

# What does the las argument do?  Look it up with ?par().
barplot(cnts, 
        horiz = TRUE, 
        col   = 'lightblue',
        las   = 2,
        cex.names = 0.5)

# Save the original graph parameters.
opar <- par(no.readonly = TRUE)

# If necessary, adjust graph parameters to make the bar plot look good then rerun barplot code.
par(mai = c(.5, 1.5, .5, .5))

par(opar)

```

## The Grouped & Stacked Barplot Challenge

```{r group_barplot}

crime <- read.csv("drink_crime.csv", stringsAsFactors = FALSE)

barplot(as.matrix(crime[, c(2:4)]), 
        legend = crime[, 1],
           col = c("red","blue","green","yellow"))

barplot(as.matrix(crime[, c(2:4)]), 
        legend = crime[, 1],
           col = c("red","blue","green","yellow"),
        beside = TRUE)

lgnd        <- crime$Crime
binge       <- crime$Binge
occassional <- crime$Occasional
never       <- crime$Never
cnts        <- data.frame(binge, occassional, never)

barplot(as.matrix(cnts),
        legend = lgnd,
        col = c("red","blue","green","yellow"))

barplot(as.matrix(cnts),
        legend = lgnd,
        col = c("red","blue","green","yellow"),
        beside = TRUE )

```


## The Pie Chart Challenge

And now for the pie chart challenge...

```{r pie}
# You can write a select statement with new column names.
tmp <- sqldf("select q3 as department, count(q3) as total from survey where q3 != ' ' group by q3")

# Or, you can execute a simple select and rename the variables afterwards.
tmp <- sqldf("select q3, count(q3) from survey where q3 != ' ' group by q3")

# Rename the columns -- department & department count.
colnames(tmp)  <- c("dept","dept_cnt")

# Add new column for academic group (HUM Humanities, SCI Sciences, SOC, Social Sciences, MED Medical)
tmp$acad_grp <- ""                       

# Set the new acad_grp column, using fix().  What happens if you use edit()?
fix(tmp)                                   

# Save a backup copy of the modified dataframe.
write.csv(tmp, file = "tmp.csv")          

# Create a new dataframe of the two columns needed for the pie chart. 
temp   <- data.frame(tmp$acad_grp, tmp$dept_cnt, stringsAsFactors = FALSE)

colnames(temp) <- c("acad_grp","dept_cnt")

grp_by <- group_by(temp, acad_grp)                    # Group rows by academic group.

slices <- summarize(grp_by, grp_tot = sum(dept_cnt))  # Sum dept_cnt for each academic group.

pie(slices$grp_tot, labels = slices$acad_grp, main = "Pie Chart") 

# We can accomplish the same thing with the pipe operator.
slices <- group_by(temp, acad_grp) %>% summarize(grp_tot = sum(dept_cnt))

pie(slices$grp_tot, labels = slices$acad_grp, main = "Pie Chart (Pipe)") 


```


## The histogram challenge

Code for the histogram challenge:

```{r histogram}

opar <- par(no.readonly = TRUE)

par(mfrow = c(3, 1))

tmp <- read.delim("home_ownership.txt", sep = "\t", stringsAsFactors = FALSE)

hist(tmp$pct_1985, 
     main = "1985", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

hist(tmp$pct_1996, 
     main = "1996", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

hist(tmp$pct_2002, 
     main = "2002", 
     col  = "lightblue",
     ylim = c(0, 27),
     xlab = "Home Ownership %")

par(opar)


```


## The box plot challenge

Code for the box plot challenge:

```{r boxplot}


```

## The scatter plot challenge

Code for the dot plot challenge:

Use the Money Supply table (3.41) from Ott (p. 137)

```{r dotplot}



```

## The stem-and-leaf Challenge

Code for the stem-and-leaf challenge:

```{r stemleaf}
 


```


